Sperm should evolve to make female meiosis fair.

### 2014-07-18 writeup_and_notes/GeneticsResub/male_dependent_drive.tex
Changes to discussion

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations
	likely precludes this interaction in plants). 
\gc{It is becoming
        increasingly clear that sperm bring a wide variety of RNA and proteins into the
        egg REF. Some of these have known functions, for example, in most
        animal species sperm-- not eggs-- are
        responsible for the transmission of the centrosome, a vitial
        component of the mitotic machinary for the zygote. 
Detailed functional studies and paternal effect mutations in model
systems further highlight
 that sperm-transmitted products have a wide-range of functions in
 the completion of syngamy, zygotic development, and the resumption of
 female meiosis \cite{Miller:01}. 
However,
        the function of many of the products the sperm brings into the
        egg is completely unknown and 
      these products vary widely over species. 
It seems quite plausible that sperm based products and
henece alleles could potentially have functional effects on
various aspects of female meiosis that occur after fertilization. } 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for \gc{conflict caused by
the such selfish systems.} Here we ask if sperm do indeed evolve to collaborate with female drivers to exploit this apparent weakness in the defense against meiotic drive. \newline 
%Why then is the requirement of fertilization to complete female meiosis so ubiquitous? 
\gc{THERE may be too much bio. detail in previous paragraph}

@@

It seems that allowing sperm to influence the outcome of female meiosis would generate a confluence of these potential conflicts -- 
	sperm could actually assist an allele that distorts female meiosis.
However, this is not the case.
We find that an allele which acts through sperm to distort female meiosis in its favor %YB [too awkward / unclear?] 
	can rarely spread through a population if it bears any cost. 
Additionally, when this self-promoting driver can spread, it can only rarely 
	be maintained as a protected polymorphism, and due to its positive frequency dependence,  
	it spends very little time at intermediate frequency.
As such, this type of exploitation cannot generate a sustained genetic conflict.
It is therefore unlikely
	that female o\"{o}genesis and meiosis will evolve to prevent their effect.  
This likely explains why females can delay the completion of meiosis until after fertilization 
	without risking exploitation by collaborations between female
        drivers and sperm alleles. \newline

@@

\yb{The theory presented herein provides a strong prediction -- in animals and other organisms in which female meiosis is not completed until after fertilization sperm will act to suppress female drive at the stage at which they can influence meiosis. 
%Therefore, for example, when fertilized by drying sperm, we expect meiosis to be fairer when  females are fertilized by 
contrary to expectations of a ``green-beard'' model, we are unsurprised by the observation that female meiosis is fairer when fertilized by sperm bearing the drive allele in two of the best described cases of female meiotic drive in animals (the \emph{Om} and \emph{In} loci in mice \citep{Agulnik1993,Wu2005}).  
Given the immense benefit to sperm of hampering female drive, drive-suppressing sperm are often likely to be fixed within a species, making the hypothesis of sperm acting drive-suppression difficult to test from intra-population crosses. 
However, crosses between populations or species are likely to provide critical tests of our theory -- specifically we predict that female meiosis will be less fair when a species (or population) is fertilized by heterospecific sperm because such sperm have not evolved to counter novel females meiotic drivers. 
We further predict that segregation in F1 females backcrossed to parental species will likely be biased with a deficit of transmission of the paternal species allele from the F1  female.   
We note that while these predictions follow straightforwardly from the theory presented above, tests of meiotic drive, and especially sperm dependent meiotic drive require a high standard of evidence to exclude plausible alternative hypotheses such as genotypic inviability including epistatic maternal X zygotic lethality.}  \newline

@@

Our results highlight potentially counterintuitive results of complex genetic conflicts. 
Despite much opportunity for conflict between sperm and females over fertilization \citep{Partridge1998},
	the interests of fertilizing sperm and female are quite well aligned during syngamy.
While conflict between mother and her alternative chromosomes ensues, 
	fertilizing sperm decidedly side with mom, as both have a shared interest in producing a viable and potentially fit offspring. 
\yb{Our model does not directly speak to the evolutionary origin of female meiotic arrest (for a review and evaluation of such hypotheses see \citep{Mira1998}); however, our work does imply that the fertilization requirement does not leave female meiosis open to sabotage by sperm collaborators, but rather that  an evolutionary loss of female meiotic arrest will result in the  loss of sperm defense against female meiotic drivers. }

>>>>

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations
	likely precludes this interaction in plants). 
\gc{It is becoming
        increasingly clear that sperm bring a wide variety of RNA and proteins into the
        egg REF. Some of these have known functions, for example, in most
        animal species sperm-- not eggs-- are
        responsible for the transmission of the centrosome, a vitial
        component of the mitotic machinary for the zygote. 
Detailed functional studies and paternal effect mutations in model
systems further highlight
 that sperm-transmitted products have a wide-range of functions in
 the completion of syngamy, zygotic development, and the resumption of
 female meiosis \cite{Miller:01, McNally:05,REF}. 
However,
        the function of many of the products the sperm brings into the
        egg is completely unknown and 
      these products vary widely over species. 
It seems quite plausible that sperm based products and
henece alleles could potentially have functional effects on
various aspects of female meiosis that occur after fertilization. } 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for \gc{conflict caused by
the such selfish systems.} Here we ask if sperm do indeed evolve to collaborate with female drivers to exploit this apparent weakness in the defense against meiotic drive. \newline 
%Why then is the requirement of fertilization to complete female meiosis so ubiquitous? 
\gc{THERE may be too much bio. detail in previous paragraph}

@@

It seems that allowing sperm to influence the outcome of female meiosis would generate a confluence of these potential conflicts -- 
	sperm could actually assist an allele that distorts female meiosis.
However, this is not the case.
We find that an allele which acts through sperm to distort female meiosis in its favor %YB [too awkward / unclear?] 
	can rarely spread through a population if it bears any cost. 
Additionally, when this self-promoting driver can spread, it can only rarely 
	be maintained as a protected polymorphism, and due to its positive frequency dependence,  
	it spends very little time at intermediate frequency.
As such, this type of exploitation cannot generate a sustained genetic conflict.
It is therefore unlikely
	that female o\"{o}genesis and meiosis will evolve to prevent their effect.  
\gc{Thus females can} delay the completion of meiosis until after fertilization 
	without risking exploitation by collaborations between female
        drivers and sperm alleles. \newline 
%%G: felt like this made it sound less why they do it.

@@

\yb{The theory presented herein provides a strong prediction -- 
organisms in which female 
meiosis is not completed until after fertilization sperm will act to suppress female drive at the stage at which they can influence meiosis. 
%Therefore, for example, when fertilized by drying sperm, we expect meiosis to be fairer when  females are fertilized by 
Contrary to expectations of a \gc{naive verbal} ``green-beard'' model, 
we are unsurprised by the observation that female meiosis is fairer when fertilized by sperm bearing the drive allele in two of the best described cases of female meiotic drive in animals (the \emph{Om} and \emph{In} loci in mice \citep{Agulnik1993,Wu2005}).  
Given the immense benefit to sperm of hampering female drive, 
drive-suppressing sperm are often likely to be fixed within a species, 
making the hypothesis of sperm acting drive-suppression difficult to test from intra-population crosses. 
However, crosses between populations or species are likely to provide critical tests of our theory -- specifically we predict that female meiosis will be less fair when a species (or population) is fertilized by heterospecific sperm because such sperm have not evolved to counter novel females meiotic drivers. 
We further predict that segregation in F1 females backcrossed to parental species will likely be biased with a deficit of transmission of the paternal species allele from the F1  female.   
We note that while these predictions follow straightforwardly from the
theory presented above, tests of meiotic drive, 
and especially sperm dependent meiotic drive require a high standard
of evidence 
to exclude plausible alternative hypotheses such as genotypic inviability including epistatic maternal X zygotic lethality.}  \newline

\gc{Our theory also encourages phylogenetic hypotheses concerning the relationship between the opportunity 
	for female meiotic drive and the requirement of fertilization for the completion of female meiosis. 
For example, we predict that a lower opportunity for female meiotic drive,
        e.g. an animal lineage with a history of high inbreeding
        or selfing, 
        may be accompanied by a relaxation of the requirement of fertilization for the
        completion of female meiosis (although this effect may be
        weaks as such
        lineages may only persist for a short time). Another
        hypothesis that our work suggests is we might
        expect that phylogenetic variation in the stage of female meiosis when
        fertilization occurs to play a role in the prevalance of
        female meiotic drive.
For example, in taxa where meiosis I is completed
by the egg pre-fertilization, including humans, centromere drive may
be more common as sperm based modifiers can not intercede than in
species where fertilization occurs during MI. 
Such hypotheses could be tested if the speed of centromere
turnover serves as a proxy for the efficacy of MI drive. }

@@

Our results highlight potentially counterintuitive results of complex genetic conflicts. 
Despite much opportunity for conflict between sperm and females over fertilization \citep{Partridge1998},
	the interests of fertilizing sperm and female are quite well aligned during syngamy.
While conflict between mother and her alternative chromosomes ensues, 
	fertilizing sperm decidedly side with mom, as both have a shared interest in producing a viable and potentially fit offspring. 
\yb{Our model does not directly speak to the evolutionary origin of
  female meiotic arrest \citep[(for a review and evaluation of such
  hypotheses see][]{Mira1998}.
However, given that arrest does occur its timing relative to
fertilisation, various features of the interaction between sperm and
egg and divergence of these features across species  \citep{Karrbook:09} could be subject
to selection to conflict between female meiotic drivers and alleles
elsewhere in the genomes of sperm and eggs.
the mechanism by which it occur
mechanistic features, and evolutionary changes among taxa

various features of the
mechanism of fertilization 
however, our work does imply that the fertilization requirement does
not 
leave female meiosis open to sabotage by sperm collaborators, but
rather that  an 
evolutionary loss of female meiotic arrest will result in the loss of sperm defense against female meiotic drivers. }
